---
title: "Vignette, ASCA"
output: 
  pdf_document: 
    keep_tex: true
vignette: >
  %\VignetteIndexEntry{Vignette, ASCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6, 
  fig.height=4
)
# Legge denne i YAML på toppen for å skrive ut til tex
#output: 
#  pdf_document: 
#    keep_tex: true
# Original:
#  rmarkdown::html_vignette:
#    toc: true
```

```{r setup}
library(multiblock)
```

# ANOVA Simultaneous Component Analysis -- ASCA

The following example uses a simulated dataset for showcasing some of the possibilities of the ASCA method.

## Simulated data

Two categorical factors and a covariate are simulated together with a standard normal set of 10 responses.

```{r cars}
set.seed(1)
dataset   <- data.frame(y=I(matrix(rnorm(24*10),ncol=10)), 
                        x=factor(c(rep(2,8),rep(1,8),rep(0,8))), 
                        z=factor(rep(c(1,0),12)), w=rnorm(24))
colnames(dataset$y) <- paste('Var', 1:10, sep=" ")
rownames(dataset)   <- paste('Obj', 1:24, sep=" ")
str(dataset)
```

## ASCA modelling

A basic ASCA model having two factors is fitted and printed.

```{r}
mod <- asca(y~x+z, data=dataset)
print(mod)
```

## Scores

Scores for first factor are extracted and a scoreplot with confidence ellipsoids is produced.

```{r}
sc <- scores(mod)
head(sc)

scoreplot(mod, legendpos = "topleft", ellipsoids = "confidence")
```

This is repeated for the second factor.

```{r}
sc <- scores(mod, factor="z")
head(sc)

scoreplot(mod, factor="z", ellipsoids = "confidence")
```

## Loadings

A basic loadingplot for the first factor is generated using graphics from the _pls_ package.

```{r}
lo <- loadings(mod)
head(lo)

loadingplot(mod, scatter = TRUE, labels = 'names')
```

