---
title: "Vignette, basic analyses"
output: 
  pdf_document: 
    keep_tex: true
vignette: >
  %\VignetteIndexEntry{Vignette, basic analyses}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6, 
  fig.height=4
)
# Legge denne i YAML på toppen for å skrive ut til tex
#output: 
#  pdf_document: 
#    keep_tex: true
# Original:
#  rmarkdown::html_vignette:
#    toc: true
```

```{r setup}
# Start the multiblock R package
library(multiblock)
```

# Basic methods

The following single- and two-block methods are available in the _multiblock_ package (function names in parentheses):

* PCA - Principal Component Analysis (_pca_)
* PCR - Principal Component Regression (_pcr_)
* PLSR - Partial Least Squares Regression (_plsr_)
* CCA - Canonical Correlation Analysis (_cca_)
* IFA - Interbattery Factor Analysis (_ifa_)
* GSVD - Generalized SVD (_gsvd_)

The following sections will describe how to format your data for analysis and invoke all methods from the list above.

## Prepare data

We use a selection of extracts from the potato data included in the package for the basic data analyses. The
data set is stored as a named list of matrices with chemical, rheological, spectral and sensory measurements.

```{r}
data(potato)
X <- potato$Chemical
y <- potato$Sensory[,1,drop=FALSE]
```

## Modelling

Since the basic methods cover both single block analysis, supervised and unsupervised analysis, the interfaces 
for the basic methods vary a bit. Supervised methods use the formula interface and the remaining methods take
input as a single matrix or list of matrices. See vignettes for supervised and unsupervised analysis for details.

```{r}
# Single block
pot.pca  <- pca(X, ncomp = 2)

# Two blocks, supervised
pot.pcr  <- pcr(y ~ X, ncomp = 2)
pot.pls  <- plsr(y ~ X, ncomp = 2)

# Two blocks, unsupervised
pot.cca  <- cca(potato[1:2])
pot.ifa  <- ifa(potato[1:2])

# Variable linked decomposition
pot.gsvd <- gsvd(lapply(potato[3:4], t))
```

## Common output elements across methods

Output from all methods include slots called _loadings_, _scores_, _blockLoadings_
and _blockScores_, or a suitable subset of these. An _info_ slot describes which
types of (block) loadings/scores are in the output. There may be various extra
elements in addition to the common elements, e.g. coefficients, weights etc.

```{r}
# PCA returns loadings and scores:
names(pot.pca)
summary(pot.pca)
# GSVD returns block scores and common loadings:
names(pot.gsvd)
summary(pot.gsvd)
```

## Scores and loadings

Functions for accessing scores and loadings are based on functions
from the _pls_ package, but extended with a _block_ parameter to
allow extraction of common/global scores/loadings and their block
counterparts. The default block is 0, corresponding to the common/global
block. Block scores/loadings can be accessed by number or name.

```{r}
# Global scores plotted with object labels
plot(scores(pot.pca), labels = "names")
```

```{r}
# Block loadings for Chemical block with variable labels in scatter format
plot(loadings(pot.cca, "Chemical"), labels = "names", scatter = TRUE)
```


```{r}
# Non-existing elements are swapped with existing ones with a warning.
sc <- scores(pot.cca)
```
